---
title: "Choosing carryover parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing carryover parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

Decay and lag have to come from somewhere. Priors and vendor benchmarks are one
answer. Cross-validation is another, and it has the advantage of being
checkable: if the procedure works, it should recover a decay you planted in
simulated data.

This vignette is about doing that correctly, because there is a tempting way to
do it that does not work.

```{r}
library(mediamix)
```

## The objective that does not work

Here is a plausible-sounding procedure. Adstock the media series, fit a linear
trend to the transformed series, extrapolate it one period ahead, and score it
against the raw series. Grid-search decay and lag; take the best.

It sounds reasonable and it is broken, for a reason that is easy to miss:
**there is no dependent variable anywhere in that loop.** The criterion never
sees the KPI. It scores how *smooth and trend-like* the transformed media
series is, which is a property of the filter, not evidence about the media.

Watch what it does on data where the true decay is 0.7:

```{r}
set.seed(42)
n <- 156
spend <- pmax(0, rnorm(n, 500, 250))
truth <- 0.7
kpi <- 200 + 0.5 * adstock_geometric(spend, decay = truth) + rnorm(n, sd = 5)

no_kpi_objective <- function(x, max_lag, decay) {
  z <- adstock_geometric(x, decay, max_lag = max_lag)
  errs <- vapply(78:(n - 1), function(e) {
    m <- lm(z[1:e] ~ seq_len(e))
    (x[e + 1] - (coef(m)[1] + coef(m)[2] * (e + 1)))^2
  }, numeric(1))
  sqrt(mean(errs))
}

g <- expand.grid(max_lag = c(2, 5, 10, 20), decay = c(0.1, 0.5, 0.9))
g$error <- mapply(function(l, d) no_kpi_objective(spend, l, d),
                  g$max_lag, g$decay)
g[order(g$error), ]
```

Two things are wrong here. The chosen decay is 0.1 against a truth of 0.7. And
look at how little separates the contenders:

```{r}
c(spread_across_whole_grid = max(g$error) / min(g$error) - 1,
  spread_across_best_four = sort(g$error)[4] / min(g$error) - 1)
```

The four best cells differ in the fifth significant figure, and the whole grid
spans under 4%. There is almost no signal to select on, and what signal there
is runs in a single direction: error rises monotonically with decay, because a
harder-smoothed series is a worse predictor of a noisy raw one. The criterion
is ranking filters by how much they smooth. It never sees the KPI, so it
cannot rank them by anything else.

## Putting the KPI in the loop

`tune_carryover()` scores each candidate `(max_lag, decay)` by how well a model
fitted on the adstocked media predicts *held-out KPI*:

```{r}
tuned <- tune_carryover(
  spend, kpi,
  max_lags = Inf,
  decays = seq(0.05, 0.95, by = 0.05)
)
tuned
```

Recovered exactly. And it is not a fluke of one decay value:

```{r}
recover <- function(true_decay, seed = 1) {
  set.seed(seed)
  x <- pmax(0, rnorm(n, 500, 250))
  y <- 200 + 0.5 * adstock_geometric(x, decay = true_decay) + rnorm(n, sd = 5)
  t <- suppressWarnings(tune_carryover(
    x, y, max_lags = Inf, decays = seq(0.05, 0.95, by = 0.05)))
  c(true = true_decay, recovered = t$best$decay,
    true_hl = half_life(true_decay), recovered_hl = t$best$half_life)
}

round(t(vapply(c(0.2, 0.35, 0.5, 0.7, 0.85), recover, numeric(4))), 3)
```

This recovery test lives in the package's own test suite. It is the check that
the objective is measuring what it claims to.

## Why slicing a fully adstocked series is legitimate

`tune_carryover()` adstocks the whole series once per grid point and then cuts
it into folds, rather than re-adstocking inside each fold. That looks like
leakage and is not, because the filter is strictly causal:

```{r}
z <- adstock_geometric(spend, decay = 0.7)
all.equal(z[1:40], adstock_geometric(spend[1:40], decay = 0.7))
all.equal(z[1:100], adstock_geometric(spend[1:100], decay = 0.7))
```

The adstock value at time *t* depends only on media at times ≤ *t*. So an
assessment period's regressor draws on training-period *media* — information
genuinely available at prediction time — and never on future media. No KPI
information crosses the boundary at any point. The package tests this
invariant explicitly, precisely because the shortcut depends on it.

## A correct objective is not the same as an easy problem

Everything above uses a single channel with no confounders. Real data is not
like that, and it is worth seeing how much harder it gets before trusting a
number.

`mm_weekly` has five correlated, flighted channels, a price term, a trend and
seasonality, and 156 weekly observations. Tune each channel's decay one at a
time against raw revenue, and:

```{r}
data(mm_weekly)
north <- mm_weekly[mm_weekly$geo == "north", ]
truth_params <- attr(mm_weekly, "truth")
grid <- seq(0.1, 0.9, by = 0.05)

channels <- names(truth_params$decay)

one_at_a_time <- vapply(channels, function(ch) {
  suppressWarnings(tune_carryover(north[[ch]], north$revenue,
                                  max_lags = Inf, decays = grid))$best$decay
}, numeric(1))

rbind(true = truth_params$decay[channels], recovered = one_at_a_time)
```

Some are close, some are badly off. Nothing is wrong with the objective — it
recovered five out of five true decays a moment ago. The problem is the
*model* it is being evaluated against: a univariate regression of revenue on
one adstocked channel, with no controls and no saturation. That model attributes
the trend, the price effect, seasonality and the other four channels' work to
whichever channel it is currently looking at, and the carryover that best
predicts under those conditions is not the true one.

Residualising the KPI against the controls first helps, but not enough:

```{r}
base <- lm(revenue ~ seq_len(nrow(north)) + price + seasonality + holiday,
           data = north)

residualised <- vapply(channels, function(ch) {
  suppressWarnings(tune_carryover(north[[ch]], residuals(base),
                                  max_lags = Inf, decays = grid))$best$decay
}, numeric(1))

rbind(true = truth_params$decay[channels],
      raw_kpi = one_at_a_time,
      residualised = residualised)
```

The lesson is not that cross-validation fails. It is that **carryover is not
separable from the rest of the specification**, so tuning it one channel at a
time against a model you do not intend to fit answers a question you did not
ask. Two ways forward:

- Pass the model you actually mean, via `fit_fn` and `predict_fn`. Note that
  these receive only the adstocked media vector and `y`, so controls have to
  reach them another way — a closure over pre-residualised `y`, as above, or
  variables captured from the enclosing environment.
- Better, tune all the channels' carryover and saturation *jointly*, inside the
  same resampling as the model itself. That is what the recipes layer is for;
  see `vignette("tidymodels")`.

Either way, `tune_carryover()` on a single channel is a diagnostic and a
starting point, not a deliverable.

## Reading the result

```{r}
head(tuned$results, 5)
```

The `results` grid is worth looking at rather than just taking `best`. If the
metric is nearly flat across a wide band of decay, the data does not identify
carryover tightly and a confident point estimate would be overstating things.

`tune_carryover()` warns when the winner is pinned to the edge of the grid:

```{r}
narrow <- tune_carryover(spend, kpi, max_lags = Inf,
                         decays = seq(0.1, 0.5, by = 0.1))
narrow$best$decay
```

Here the truth is 0.7 and the grid stops at 0.5, so the optimum is outside it.
The warning is the signal to widen the grid rather than to believe the answer.

## Sanity-check the answer against the data you have

```{r}
window <- effective_window(tuned$best$decay, coverage = 0.90)
window
```

The selected decay of `r round(tuned$best$decay, 2)` implies `r window` periods
before 90% of a burst's effect has played out. Compare that against the series
length: if the effective window approaches a large fraction of it, the
parameter is not identified by the data no matter how confident the
cross-validation looks. Here `r window` periods against `r n` is comfortable.

## Resampling schemes

Both schemes are forward-only; neither ever trains on data that follows what it
assesses.

```{r}
rolling <- tune_carryover(spend, kpi, max_lags = Inf,
                          decays = seq(0.5, 0.9, by = 0.1),
                          scheme = "rolling_origin")
kfold <- tune_carryover(spend, kpi, max_lags = Inf,
                        decays = seq(0.5, 0.9, by = 0.1),
                        scheme = "k_fold_forward", k = 5)
c(rolling = rolling$best$decay, k_fold = kfold$best$decay)
```

`"rolling_origin"` grows the training window one period at a time and gives
many splits; `"k_fold_forward"` uses a handful of contiguous forward blocks and
is cheaper on long series.

For tuning carryover jointly with saturation and a model's own penalty, see
`vignette("tidymodels")`.
