---
title: "Choosing carryover parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing carryover parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, out.width = "100%", dpi = 110)
options(digits = 4)
```

Decay and lag have to come from somewhere. Priors and vendor benchmarks are one
answer. Cross-validation is another, and it has the advantage of being
checkable: if the procedure works, it should recover a decay you planted in
simulated data.

This vignette is about doing that correctly, because there is a tempting way to
do it that does not work.

```{r}
library(mediamix)
```

## The objective that does not work

Here is a plausible-sounding procedure. Adstock the media series, fit a linear
trend to the transformed series, extrapolate it one period ahead, and score it
against the raw series. Grid-search decay and lag; take the best.

It sounds reasonable and it is broken, for a reason that is easy to miss:
**there is no dependent variable anywhere in that loop.** The criterion never
sees the KPI. It scores how *smooth and trend-like* the transformed media
series is, which is a property of the filter, not evidence about the media.

Watch what it does on data where the true decay is 0.7:

```{r}
set.seed(42)
n <- 156
spend <- pmax(0, rnorm(n, 500, 250))
truth <- 0.7
kpi <- 200 + 0.5 * adstock_geometric(spend, decay = truth) + rnorm(n, sd = 5)

no_kpi_objective <- function(x, max_lag, decay) {
  z <- adstock_geometric(x, decay, max_lag = max_lag)
  errs <- vapply(78:(n - 1), function(e) {
    m <- lm(z[1:e] ~ seq_len(e))
    (x[e + 1] - (coef(m)[1] + coef(m)[2] * (e + 1)))^2
  }, numeric(1))
  sqrt(mean(errs))
}

g <- expand.grid(max_lag = c(2, 5, 10, 20), decay = c(0.1, 0.5, 0.9))
g$error <- mapply(function(l, d) no_kpi_objective(spend, l, d),
                  g$max_lag, g$decay)
g[order(g$error), ]
```

Two things are wrong here. The chosen decay is 0.1 against a truth of 0.7. And
look at how little separates the contenders:

```{r}
c(spread_across_whole_grid = max(g$error) / min(g$error) - 1,
  spread_across_best_four = sort(g$error)[4] / min(g$error) - 1)
```

The four best cells differ in the fifth significant figure, and the whole grid
spans under 4%. There is almost no signal to select on, and what signal there
is runs in a single direction: error rises monotonically with decay, because a
harder-smoothed series is a worse predictor of a noisy raw one. The criterion
is ranking filters by how much they smooth. It never sees the KPI, so it
cannot rank them by anything else.

## Putting the KPI in the loop

`tune_carryover()` scores each candidate `(max_lag, decay)` by how well a model
fitted on the adstocked media predicts *held-out KPI*:

```{r}
tuned <- tune_carryover(
  spend, kpi,
  max_lags = Inf,
  decays = seq(0.05, 0.95, by = 0.05)
)
tuned
```

Recovered exactly. The profile shows how sharply the data identifies it —
a clear valley here, which is what a well-identified carryover looks like:

```{r}
plot(tuned, truth = truth)
```

And it is not a fluke of one decay value:

```{r}
recover <- function(true_decay, seed = 1) {
  set.seed(seed)
  x <- pmax(0, rnorm(n, 500, 250))
  y <- 200 + 0.5 * adstock_geometric(x, decay = true_decay) + rnorm(n, sd = 5)
  t <- suppressWarnings(tune_carryover(
    x, y, max_lags = Inf, decays = seq(0.05, 0.95, by = 0.05)))
  c(true = true_decay, recovered = t$best$decay,
    true_hl = half_life(true_decay), recovered_hl = t$best$half_life)
}

round(t(vapply(c(0.2, 0.35, 0.5, 0.7, 0.85), recover, numeric(4))), 3)
```

This recovery test lives in the package's own test suite. It is the check that
the objective is measuring what it claims to.

## Why slicing a fully adstocked series is legitimate

`tune_carryover()` adstocks the whole series once per grid point and then cuts
it into folds, rather than re-adstocking inside each fold. That looks like
leakage and is not, because the filter is strictly causal:

```{r}
z <- adstock_geometric(spend, decay = 0.7)
all.equal(z[1:40], adstock_geometric(spend[1:40], decay = 0.7))
all.equal(z[1:100], adstock_geometric(spend[1:100], decay = 0.7))
```

The adstock value at time *t* depends only on media at times ≤ *t*. So an
assessment period's regressor draws on training-period *media* — information
genuinely available at prediction time — and never on future media. No KPI
information crosses the boundary at any point. The package tests this
invariant explicitly, precisely because the shortcut depends on it.

## A correct objective is not the same as an easy problem

Everything above uses a single channel with no confounders. Real data is not
like that, and it is worth seeing how much harder it gets before trusting a
number.

`mm_weekly` has five correlated, flighted channels, a price term, a trend and
seasonality, and 156 weekly observations. Tune each channel's decay one at a
time against raw revenue, and:

```{r}
data(mm_weekly)
north <- mm_weekly[mm_weekly$geo == "north", ]
truth_params <- attr(mm_weekly, "truth")
grid <- seq(0.1, 0.9, by = 0.05)

channels <- names(truth_params$decay)

one_at_a_time <- vapply(channels, function(ch) {
  suppressWarnings(tune_carryover(north[[ch]], north$revenue,
                                  max_lags = Inf, decays = grid))$best$decay
}, numeric(1))

rbind(true = truth_params$decay[channels], recovered = one_at_a_time)
```

Some are close, some are badly off. Nothing is wrong with the objective — it
recovered five out of five true decays a moment ago. The problem is the
*model* it is being evaluated against: a univariate regression of revenue on
one adstocked channel, with no controls and no saturation. That model attributes
the trend, the price effect, seasonality and the other four channels' work to
whichever channel it is currently looking at, and the carryover that best
predicts under those conditions is not the true one.

Residualising the KPI against the controls first helps, but not enough:

```{r}
base <- lm(revenue ~ seq_len(nrow(north)) + price + seasonality + holiday,
           data = north)

residualised <- vapply(channels, function(ch) {
  suppressWarnings(tune_carryover(north[[ch]], residuals(base),
                                  max_lags = Inf, decays = grid))$best$decay
}, numeric(1))

rbind(true = truth_params$decay[channels],
      raw_kpi = one_at_a_time,
      residualised = residualised)
```

The lesson is not that cross-validation fails. It is that **carryover is not
separable from the rest of the specification**, so tuning it one channel at a
time against a model you do not intend to fit answers a question you did not
ask.

## Tuning every channel at once

`tune_carryover_joint()` puts all five channels and the controls into one
model and chooses each channel's decay with the others' effects accounted for.
It searches by coordinate descent — one channel at a time, holding the rest,
repeating until nothing moves — so it costs a few sweeps of the one-channel
grid rather than the full five-dimensional grid:

```{r}
controls <- data.frame(week = seq_len(nrow(north)), price = north$price,
                       seasonality = north$seasonality,
                       holiday = north$holiday)

joint <- tune_carryover_joint(north[channels], north$revenue,
                              controls = controls, decays = grid)
joint

rbind(true = truth_params$decay[channels],
      one_at_a_time = one_at_a_time,
      residualised = residualised,
      joint = joint$best$decay,
      joint_1se = joint$best_1se$decay)
```

Much closer, and the profiles show where the remaining doubt lies:

```{r fig.height = 5.5}
plot(joint, truth = truth_params$decay)
```

Television, video and social have clear valleys. Search's profile barely
moves across the whole range: the data cannot tell a search half-life of a
fraction of a week from one of several, and the one-standard-error choice
falls back to the short end, near the truth. That flatness is itself the
finding, and a point estimate without the profile would hide it.

The model here is still linear in adstocked media, with no saturation, so it
is not the model the data came from. For carryover tuned jointly with
saturation and a model penalty, use the recipe steps in
`vignette("tidymodels")`.

## Reading the result

```{r}
head(tuned$results, 5)
```

The `results` grid is worth looking at rather than just taking `best`. If the
metric is nearly flat across a wide band of decay, the data does not identify
carryover tightly and a confident point estimate would be overstating things.

Two things in the output help with that. `std_err` is the standard error of
the per-split errors, which says how far apart two candidates need to be before
the difference means anything. And `best_1se` applies the
one-standard-error rule (Breiman et al., 1984; Hastie, Tibshirani and
Friedman, 2009): the *shortest* carryover whose error is within one standard
error of the minimum.

```{r}
rbind(best = tuned$best, one_se = tuned$best_1se)
```

When the two agree, the data has spoken clearly. When they do not, the data
cannot tell them apart, and the shorter carryover is the more conservative
claim to put in front of a client — a long carryover flatters a channel by
crediting it with sales weeks after it ran.

By default the metric is *pooled*: every out-of-sample forecast across all
splits is collected and `metric_fn` is evaluated once, so `rmse` is the root
mean squared forecast error. `aggregate = "mean"` averages a per-split metric
instead, which is the `tune` convention; with one-period assessment sets that
turns any RMSE into a mean absolute error, because the RMSE of a single point
is its absolute error.

`tune_carryover()` warns when the winner is pinned to the edge of the grid:

```{r}
narrow <- tune_carryover(spend, kpi, max_lags = Inf,
                         decays = seq(0.1, 0.5, by = 0.1))
narrow$best$decay
```

Here the truth is 0.7 and the grid stops at 0.5, so the optimum is outside it.
The warning is the signal to widen the grid rather than to believe the answer.

## The start of the series

Adstock starts from zero unless told otherwise, as if no media had ever run
before the first week. For a slow channel that understates the early weeks:

```{r}
steady <- rep(100, 8)
rbind(cold = adstock_geometric(steady, decay = 0.85),
      warm = adstock_geometric(steady, decay = 0.85,
                               state = adstock_steady_state(steady, 0.85)))
```

`adstock_steady_state()` seeds the filter as though the early average spend
had always been running, and `warm_start = TRUE` applies it to every
candidate in `tune_carryover()` and `tune_carryover_joint()`. Real pre-period
spend, passed through `adstock_state()`, is better still when you have it.

## Sanity-check the answer against the data you have

```{r}
window <- effective_window(tuned$best$decay, coverage = 0.90)
window
```

The selected decay of `r round(tuned$best$decay, 2)` implies `r window` periods
before 90% of a burst's effect has played out. Compare that against the series
length: if the effective window approaches a large fraction of it, the
parameter is not identified by the data no matter how confident the
cross-validation looks. Here `r window` periods against `r n` is comfortable.

## Resampling schemes

Both schemes are forward-only; neither ever trains on data that follows what it
assesses.

```{r}
rolling <- tune_carryover(spend, kpi, max_lags = Inf,
                          decays = seq(0.5, 0.9, by = 0.1),
                          scheme = "rolling_origin")
kfold <- tune_carryover(spend, kpi, max_lags = Inf,
                        decays = seq(0.5, 0.9, by = 0.1),
                        scheme = "k_fold_forward", k = 5)
c(rolling = rolling$best$decay, k_fold = kfold$best$decay)
```

`"rolling_origin"` grows the training window one period at a time and gives
many splits; `"k_fold_forward"` uses a handful of contiguous forward blocks and
is cheaper on long series.

For tuning carryover jointly with saturation and a model's own penalty, see
`vignette("tidymodels")`.

## References

Bergmeir, C. and Benítez, J. M. (2012). On the use of cross-validation for
time series predictor evaluation. *Information Sciences*, 191, 192–213.

Breiman, L., Friedman, J., Olshen, R. and Stone, C. (1984). *Classification and
Regression Trees*. Wadsworth.

Hastie, T., Tibshirani, R. and Friedman, J. (2009). *The Elements of
Statistical Learning* (2nd ed.). Springer.
