---
title: "Building journeys from event logs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building journeys from event logs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

Attribution packages consume journeys — `"display > social > paid_search"`
strings with conversion counts attached. Almost none of them produce those
strings. `ChannelAttribution` has consumed them for eleven years and has never
shipped the step that builds them, on the reasonable grounds that its job is
Markov removal effects and it does that in C++.

So the step gets done in analysis code, usually as a `group_by()` and a
`paste(collapse = " > ")`. This vignette is about what that loses, because the
answer is not small and it feeds the transition matrix directly.

```{r}
library(mediamix)
data(mm_events)

nrow(mm_events)
head(mm_events)
```

`mm_events` is a synthetic log of about 21,000 touchpoint events across 5,400
customers. It contains, on purpose, everything a real export contains: repeat
converters, consecutive duplicate channels, direct and blank and missing
channel labels, customers who never convert, timestamp ties, and rows in no
particular order.

## The naive approach

One path per customer, in timestamp order:

```{r}
ordered <- mm_events[order(mm_events$customer_id, mm_events$timestamp), ]
naive <- split(ordered$channel, ordered$customer_id)

c(journeys = length(naive), mean_length = mean(lengths(naive)))
```

5,400 journeys at a mean length of 3.85. Now the same log through
`build_paths()`:

```{r}
paths <- build_paths(
  mm_events,
  id = "customer_id",
  channel = "channel",
  timestamp = "timestamp",
  conversion = "conversion",
  value = "value"
)
path_summary(paths)
```

6,137 journeys — 14% more — at a mean length of 2.82, down 27%. Journey counts
and lengths feed a Markov transition matrix directly, so a 14% error in the
count and a 27% error in the length is not a rounding difference in the final
numbers.

Here is where the difference comes from, one concern at a time.

```{r}
step <- function(...) {
  s <- path_summary(build_paths(
    mm_events, id = "customer_id", channel = "channel",
    timestamp = "timestamp", conversion = "conversion", ...))
  c(journeys = s$journeys, mean_length = round(s$mean_length, 2))
}

rbind(
  `naive`                       = c(length(naive), round(mean(lengths(naive)), 2)),
  `split on conversion`         = step(split_on = "conversion",
                                       collapse_repeats = FALSE),
  `+ 30-day inactivity gap`     = step(split_on = c("conversion", "gap"),
                                       gap = 30, collapse_repeats = FALSE),
  `+ collapse duplicates`       = step(split_on = c("conversion", "gap"),
                                       gap = 30, collapse_repeats = TRUE)
)
```

## The seven things that go wrong

### 1. Journey splitting

A customer who converts in March and again in September is two journeys. Treat
them as one and March's touchpoints take credit for September's sale, while
the journey count falls by every repeat conversion in the file.

```{r}
repeat_buyers <- table(mm_events$customer_id[mm_events$conversion == 1])
table(as.integer(repeat_buyers))
```

`split_on` also accepts `"gap"`, so a customer who disappears for four months
and comes back starts a new journey rather than extending an eleven-month one.

### 2. Lookback windows

A touch two years before a conversion did not cause it. Thirty, sixty and
ninety days are the standard choices, and they are standard because everybody
copied each other.

`conversion_lag()` lets you read the window off your own data instead:

```{r}
conversion_lag(paths)
```

Ninety per cent of journeys in this log complete within about fourteen days, so
a 30-day window truncates almost nothing and a 90-day window is pure
decoration. The long tail is real, though — the slowest journey takes 135 days
— so a window is a decision about which tail to discard, not a free
simplification. On a considered-purchase category the same table will tell you
the opposite. No other package offers this, and it takes one line.

### 3. Non-converting journeys

These are kept by default, and you should think hard before changing that.

```{r}
with_nulls <- as_channel_paths(paths)
head(with_nulls, 4)
c(converting = sum(with_nulls$total_conversions),
  null = sum(with_nulls$total_null))
```

Markov removal effects are computed *against* the null paths: the transition
matrix needs an absorbing failure state to compare the conversion state with.
Drop them and every channel's estimated effect is biased upward, silently.

### 4. Direct, none and missing channels

Every real log has them, and every tutorial handles them differently.

```{r}
table(build_paths(mm_events, id = "customer_id", channel = "channel",
                  timestamp = "timestamp", conversion = "conversion",
                  direct = "label")$channel)
```

`"keep"` leaves the labels alone (naming the unnamed ones so they can be
referred to at all), `"label"` folds them into one `"(direct)"` channel, and
`"drop"` removes those touches entirely. The choice moves the numbers, so it
is an argument rather than a default buried in the source.

### 5. Accounting for what gets dropped

Dropping direct traffic can strip a converting journey of every touchpoint it
had — including, sometimes, the touch that recorded the conversion. mediamix
fixes the conversion facts before any filtering, so the conversion is never
erased; it is counted and reported as unattributable.

```{r}
dropped <- build_paths(
  mm_events, id = "customer_id", channel = "channel",
  timestamp = "timestamp", conversion = "conversion",
  lookback = 1, direct = "drop"
)
path_summary(dropped)[, c("conversions_observed", "converting_journeys",
                          "conversions_unattributable")]
```

Those conversions happened. They just cannot be credited to any channel under
these settings, and a pipeline that quietly reports a smaller denominator is
lying by omission.

### 6. Consecutive duplicates

Three page views on the same channel are usually one exposure recorded three
times. Collapsing them is what takes the mean journey length from 3.38 to
2.81 in the table above. `collapse_repeats = TRUE` keeps the *first* touch of each run,
preserving when the channel entered the journey — if you are using
`credit_time_decay()`, consider `FALSE`, since keeping the first of a run
pushes each channel's apparent recency backward.

### 7. Deterministic ordering

Logs written at second resolution are full of ties. `build_paths()` breaks
them by original row order, so the same input always gives the same journeys —
which matters because first-touch and last-touch results are otherwise not
reproducible.

## Assigning credit

Five rules, plus a hook for your own. Each returns the journey table with a
`credit` column, so they compose and compare.

```{r}
lin <- credit_linear(paths)
head(lin[, c("path_id", "channel", "touch_rank", "touch_n", "credit")])
```

Credit sums to 1 within every converting journey, and total credit equals the
number of attributable conversions — the invariant that catches journeys
silently dropped in construction:

```{r}
c(total_credit = sum(lin$credit),
  attributable = path_summary(paths)$converting_journeys)
```

`attribute()` runs several rules at once, which is the point:

```{r}
res <- attribute(paths)
head(res, 8)
```

## The spread is the finding

Any single rule gives you a number. Several rules give you the range that
number could have been, which is more honest and usually more useful:

```{r}
attribution_spread(res)
```

Display's share runs from 7% to 29% depending only on the convention chosen.
That channel has not been measured; it has been assigned a number. Paid search
and email move the other way. `channel_positions()` says why:

```{r}
channel_positions(paths)
```

Display opens journeys — it appears first six times for every once it appears
last — while paid search and email close them. First-touch and last-touch
attribution therefore disagree about them by construction, and reporting either
one alone is choosing a side of that disagreement without saying so.

Note that both tables carry a column called `first_last_ratio` and they are not
the same quantity. Here it is a ratio of touchpoint *counts* (6.10 for
display); in `attribution_spread()` above it is a ratio of credited *shares*
(4.06). Both say display opens journeys; they are measured in different
currencies and will not agree numerically.

The assist matrix is the other half of the picture:

```{r}
real <- c("display", "social", "organic_search", "paid_search", "email")
round(assisted_conversions(paths, normalise = TRUE)[real, real], 3)
```

Read a row as "this channel assisted these channels". Display precedes paid
search in 42% of the converting journeys display appears in; paid search
precedes display in only 10% of its own. That asymmetry is the same finding as
the first-versus-last ratio, seen pairwise rather than in aggregate.

## Time decay, and the connection to the other half

`credit_time_decay()` spreads a conversion's credit backward across prior
touchpoints with geometric decay — the same kernel `adstock_geometric()` uses
to spread spend forward, and the same `decay` vocabulary:

```{r}
td <- credit_time_decay(paths, decay = decay_from_half_life(3))
example <- td[td$path_id == td$path_id[td$converted & td$touch_n >= 4][1], ]
example[, c("channel", "touch_rank", "time_to_conversion", "credit")]
```

That symmetry is not a coincidence. See `vignette("spine")`.

## Handing off to ChannelAttribution

mediamix does not compute Markov removal effects, because `ChannelAttribution`
does it properly in C++. `as_channel_paths()` exports the journeys in exactly
the format it expects:

```{r}
ca <- as_channel_paths(paths)
head(ca, 4)
```

```{r ca, eval = requireNamespace("ChannelAttribution", quietly = TRUE)}
ChannelAttribution::markov_model(
  ca, var_path = "path",
  var_conv = "total_conversions", var_null = "total_null"
)
```

```{r eval = !requireNamespace("ChannelAttribution", quietly = TRUE), echo = FALSE, comment = NA}
cat("ChannelAttribution is not installed, so the chunk above was not run.\n")
```

That is twenty lines of interop and it makes mediamix the front end to a good
package rather than a worse competitor to it.

## One caveat, stated plainly

Rule-based attribution is a bookkeeping convention, not a causal estimate. It
divides observed conversions among observed touchpoints according to a rule you
chose. It cannot tell you what would have happened if a channel had not run,
because that outcome is not in the log — no rule, however sophisticated,
recovers a counterfactual from data that does not contain one. Only an
experiment does.

Well-built journeys make the bookkeeping honest, comparable across rules, and
reproducible. That is worth a great deal. It is not incrementality.
