---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, out.width = "100%", dpi = 110)
options(digits = 4)
```

mediamix turns raw marketing data into model-ready features. It does two
things, and deliberately not a third: it converts media spend into
carryover- and saturation-adjusted regressors, and it converts raw event logs
into attributed customer journeys. It does not fit models. `lm()`, `glmnet`
and `brms` already do that better than a marketing package would.

This vignette walks the whole arc on the media side: raw weekly spend, through
diagnostics and transforms, to a fitted model, contributions and return on
investment.

```{r}
library(mediamix)
data(mm_weekly)

north <- mm_weekly[mm_weekly$geo == "north", ]
channels <- c("tv", "video", "search", "social", "display")
str(north[, c("date", channels, "price", "revenue")])
```

## Look before you transform

Most marketing mix projects fail for reasons visible in the data before any
model is fitted. `diagnose_media()` runs those checks.

```{r}
diagnose_media(mm_weekly, media = channels, by = "geo")
```

Nothing structural is wrong here, which is what you would expect from
synthetic data. On real data this is where you learn that two channels were
planned together and their coefficients will therefore flip sign on any small
change to the specification. The detail is in the pieces:

```{r}
d <- diagnose_media(mm_weekly, media = channels, by = "geo")
d$variation
```

Read `zero_share` as the worst geography's dark-week rate, not an overall
average: with `by` supplied, every column here reports the worst case across
series, which is what you want from a diagnostic. `video` goes dark in 42% of
weeks in its worst geography and `display` in a third. That is normal — media
is flighted — but it means those channels carry less information than
156 weeks per geography suggests.

## Carryover

Advertising does not stop working the week it runs. Adstock spreads each
period's spend forward with geometric decay.

```{r}
spend <- c(100, 0, 0, 0, 0, 0)
round(adstock_geometric(spend, decay = 0.5), 2)
```

Practitioners think in half-lives, not decay coefficients, so the package
provides the translation and uses it everywhere:

```{r}
theta <- decay_from_half_life(3)   # a three-week half-life on weekly data
theta
half_life(theta)
effective_window(theta, coverage = 0.90)
```

That last line says a three-week half-life still has 10% of its effect
outstanding after 10 weeks — which is how you decide whether truncating the
kernel costs you anything, and whether 156 weeks of data can identify a
carryover this long at all.

### Normalisation changes what your coefficient means

This is the argument that most often changes an answer without anyone
noticing, so it is explicit rather than implied.

```{r}
constant <- rep(100, 50)
tail(adstock_geometric(constant, decay = 0.8), 1)                     # 100
tail(adstock_geometric(constant, decay = 0.8, normalise = FALSE), 1)  # 500
```

Normalised, the kernel sums to 1: a constant spend of 100 adstocks to 100, the
series level is preserved, and the fitted coefficient reads as the effect of
*one unit of media*. Unnormalised — raw Koyck accumulation — the same series
accumulates to `100 / (1 - 0.8)`, and the coefficient reads as the effect of
one unit of *accumulated* media. Both are defensible. Comparing a coefficient
fitted one way against a coefficient fitted the other is not.

mediamix defaults to normalised, as does the Bayesian MMM of Jin et al.
(2017). Robyn's geometric adstock does not, so if you are reconciling against
Robyn, set `normalise = FALSE`.

### Dark weeks

Media data is full of zero-spend weeks, and carryover has to survive them:

```{r}
flighted <- c(100, 100, 0, 0, 0, 0, 100)
round(adstock_geometric(flighted, decay = 0.6), 2)
```

Carryover decays across the gap rather than resetting, and the new burst adds
to what is left.

### Delayed peaks

A geometric kernel always peaks in the period of spend. Television and
out-of-home routinely do not, and no decay rate fixes that. Weibull adstock
can:

```{r}
w_geo <- adstock_weights(8, decay = 0.6)
w_wei <- adstock_weights_weibull(8, shape = 2, scale = 3, type = "pdf")
w_del <- adstock_weights_delayed(8, decay = 0.6, peak = 2)
round(rbind(geometric = w_geo, weibull = w_wei, delayed = w_del), 3)
which.max(w_wei)
```

The delayed kernel of Jin et al. (2017) does the same with a `peak` read
directly in weeks. The three shapes side by side:

```{r kernels, fig.height = 3.6}
cols <- mm_palette(3)
ink <- mm_palette(role = "ink")
op <- par(mar = c(4, 4.5, 1, 1), bg = ink[["surface"]], fg = ink[["axis"]])
matplot(0:7, cbind(w_geo, w_wei, w_del), type = "o", pch = 19, lty = 1,
        lwd = 2, col = cols, xlab = "Weeks after spend",
        ylab = "Share of the effect", las = 1, bty = "n",
        col.axis = ink[["muted"]], col.lab = ink[["secondary"]])
legend("topright", c("geometric, decay 0.6", "Weibull pdf, shape 2",
                     "delayed, peak at week 2"),
       col = cols, lwd = 2, pch = 19, bty = "n", text.col = ink[["secondary"]])
par(op)
```

## Saturation

Adstock says *when* money works. Saturation says how hard it works at the
margin.

```{r}
levels <- c(0, 25, 50, 100, 200, 400) * 1000
round(rbind(
  concave = saturate_hill(levels, half_max = 100000),
  s_shape = saturate_hill(levels, half_max = 100000, shape = 3)
), 3)
```

```{r saturation, fig.height = 3.6}
grid_spend <- seq(0, 400000, by = 2000)
op <- par(mar = c(4, 4.5, 1, 1), bg = ink[["surface"]], fg = ink[["axis"]])
plot(grid_spend / 1000, saturate_hill(grid_spend, half_max = 100000),
     type = "l", lwd = 2, col = cols[1], ylim = c(0, 1), las = 1, bty = "n",
     xlab = "Spend (thousands)", ylab = "Share of the ceiling",
     col.axis = ink[["muted"]], col.lab = ink[["secondary"]])
lines(grid_spend / 1000, saturate_hill(grid_spend, 100000, shape = 3),
      lwd = 2, col = cols[2])
abline(v = 100, col = ink[["muted"]])
text(100, 0.05, "half_max", pos = 4, cex = 0.8, col = ink[["muted"]])
legend("bottomright", c("shape = 1, concave", "shape = 3, S-shaped"),
       col = cols[1:2], lwd = 2, bty = "n", text.col = ink[["secondary"]])
par(op)
```

`half_max` is the spend at which the response reaches half its ceiling, which
is interpretable in the units you actually budget in. `shape > 1` produces the
threshold effect below which media barely registers.

## Composing them, in the right order

Saturation goes *after* adstock. Saturating first caps each period's spend in
isolation and then lets carryover accumulate the capped values past the
ceiling, which defeats the point of having a ceiling. `media_transform()`
composes them correctly and refuses to do it backwards quietly.

```{r}
truth <- attr(mm_weekly, "truth")

transformed <- as.data.frame(Map(
  function(x, decay, half_max, shape) {
    media_transform(
      x,
      adstock = list(decay = decay),
      saturation = list(half_max = half_max, shape = shape)
    )
  },
  north[channels], truth$decay[channels],
  truth$half_max[channels], truth$shape[channels]
))
round(head(transformed, 4), 4)
```

`mm_weekly` was generated from known parameters, attached as its `"truth"`
attribute, so here we can transform with the right ones. On real data you do
not know them — see `vignette("carryover")` for how to choose them.

## Fit whatever model you like

mediamix stops here. The transformed columns are ordinary numeric vectors:

```{r}
model_data <- transformed
model_data$week <- seq_len(nrow(model_data))
model_data$price <- north$price
model_data$seasonality <- north$seasonality
model_data$holiday <- north$holiday

fit <- lm(north$revenue ~ ., data = model_data)
round(coef(fit)[channels])
round(truth$beta[channels])
```

Close, not exact — 156 weekly observations of flighted, partly collinear media
is a genuinely hard estimation problem, and pretending otherwise is how
marketing mix models acquire their reputation.

The controls are doing real work. Drop them and the media coefficients absorb
the trend and the price effect instead:

```{r}
bare <- lm(north$revenue ~ ., data = transformed)
round(rbind(true = truth$beta[channels],
            with_controls = coef(fit)[channels],
            without = coef(bare)[channels]))
```

Two of the five turn negative, and `search` collapses from
`r round(coef(fit)[["search"]])` to `r round(coef(bare)[["search"]])`.
Nothing about the transform changed; the specification did.

## Contributions and ROI

A marketing mix deliverable is not the coefficients. It is the decomposition.

```{r}
contrib <- contributions(transformed, fit, index = north$date)
round(tapply(contrib$contribution, contrib$channel, sum))
```

`"(baseline)"` is everything the model predicts that this decomposition does
not attribute to the supplied media — the intercept, the trend, price,
seasonality and the December uplift, pooled. It is *not* organic demand, and
reporting it as such is the commonest way one of these decks overstates its
own precision.

Over time, with the baseline left out so the media is visible:

```{r contributions-plot}
plot(contrib)
```

The dip at the very start of 2023 is the cold start: television's long
carryover takes weeks to fill from zero. `adstock_steady_state()` removes it
when that matters — see `vignette("carryover")`.

Pooling it that way is what makes the arithmetic close exactly:

```{r}
totals <- as.numeric(tapply(contrib$contribution, contrib$period, sum))
all.equal(totals, unname(fitted(fit)))
```

Media accounts for about a fifth of predicted revenue here, which is a
plausible figure for a category with this much baseline:

```{r}
media_share <- 1 - sum(contrib$contribution[contrib$channel == "(baseline)"]) /
  sum(contrib$contribution)
round(media_share, 3)
```

```{r}
roi(contrib, north[channels])
```

### Average and marginal return are different questions

Average ROI is a scorecard. Marginal ROI is the decision variable — what the
*next* pound returns — and it is the only one of the two that should drive a
reallocation.

The tempting shortcut is to differentiate the saturation curve at the mean
adstocked level and multiply by the kernel's first weight, `1 - decay`, on the
grounds that a pound spent this week only adds that much to this week's
adstock. It does — but it also adds to every later week's adstock, and under a
normalised kernel those contributions sum to the whole pound. The shortcut
measures the return *inside the week of spend* and ignores the rest, which for
a slow channel is most of it.

`marginal_roi()` avoids the calculus altogether. It re-runs the full transform
on a budget 1% larger in every week and divides the extra response by the
extra spend — the definition used by Jin et al. (2017) and by Meridian —
so carryover and flighting are both handled by construction, and the result is
a total over the same weeks as the average:

```{r}
marginal <- do.call(rbind, lapply(channels, function(ch) {
  cbind(channel = ch, marginal_roi(
    north[[ch]], coefficient = coef(fit)[[ch]],
    adstock = list(decay = truth$decay[[ch]]),
    saturation = list(half_max = truth$half_max[[ch]],
                      shape = truth$shape[[ch]])
  ))
}))
marginal[, c("channel", "roi", "mroi")]
```

Compare the shortcut for television, whose decay of 0.85 means only 15% of a
week's spend lands in that week's adstock:

```{r}
z_tv <- adstock_geometric(north$tv, decay = truth$decay[["tv"]])
shortcut <- mroi(mean(z_tv), coefficient = coef(fit)[["tv"]],
                 half_max = truth$half_max[["tv"]],
                 shape = truth$shape[["tv"]]) * (1 - truth$decay[["tv"]])
c(shortcut = shortcut, simulated = marginal$mroi[marginal$channel == "tv"])
```

The shortcut understates television's marginal return about six-fold, and a
reallocation built on it would strip budget from exactly the channel whose
effect arrives late.

Read the simulated column instead. Display's average pound returns about 1.5,
but its *next* pound returns less than it costs: it is over-funded, and the
average-ROI table hides that completely. Search and social have average
returns roughly double their marginal ones — concave curves, well up the
slope. Television is the odd one out: its marginal return is close to its
average, because its S-shaped curve (`shape = 1.6`) is still near its steepest
section at this budget. Under a concave curve marginal is below average
everywhere; under an S-curve it can be close to or even above it below the
inflection point, and assuming otherwise is how a threshold channel stays
starved.

The full curve, and its inverse:

```{r response-curve, fig.height = 5.5}
curve <- response_curve(
  spend = seq(0, 6000, by = 1000),
  coefficient = coef(fit)[["tv"]],
  half_max = truth$half_max[["tv"]], shape = truth$shape[["tv"]]
)
curve

fine <- response_curve(seq(0, 6000, by = 50), coefficient = coef(fit)[["tv"]],
                       half_max = truth$half_max[["tv"]],
                       shape = truth$shape[["tv"]])
plot(fine, observed = adstock_geometric(north$tv, truth$decay[["tv"]]))

spend_for(target = 3000, coefficient = coef(fit)[["tv"]],
          half_max = truth$half_max[["tv"]], shape = truth$shape[["tv"]])
```

These describe the curve the model fitted, which is not the same as what would
happen if you actually spent that much. The curve is identified only over the
range of spend the data contains; anything beyond it is extrapolation wearing
a decimal point.

## Where next

- `vignette("carryover")` — choosing decay and lag by cross-validation against
  your KPI rather than by eye.
- `vignette("journeys")` — the attribution half: building customer journeys
  from raw event logs.
- `vignette("tidymodels")` — tuning carryover, saturation and model penalty
  jointly.
- `vignette("spine")` — why carryover and credit are the same idea.

## References

Jin, Y., Wang, Y., Sun, Y., Chan, D. and Koehler, J. (2017). *Bayesian
methods for media mix modeling with carryover and shape effects.* Google Inc.
