---
title: "Carryover and credit are the same idea"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Carryover and credit are the same idea}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

mediamix does two things that look unrelated. It adstocks media spend, which is
a marketing mix modelling concern. It assigns credit across customer journeys,
which is a multi-touch attribution concern. These live in different teams,
different literatures and usually different packages.

They are the same operation run in opposite directions.

```{r}
library(mediamix)
```

## Forward: one impulse of spend

Adstock takes a single burst of media and spreads its effect forward through
time with geometric decay:

```{r}
burst <- c(100, 0, 0, 0, 0, 0)
round(adstock_geometric(burst, decay = 0.5, normalise = FALSE), 2)
```

100, 50, 25, 12.5 — each period retains half of the last. The kernel is
$\theta^i$ at lag $i$.

## Backward: one conversion

Time-decay attribution takes a single conversion and spreads its credit
backward across the touchpoints that preceded it, with geometric decay:

```{r}
journey <- data.frame(
  cust = "a",
  channel = c("display", "social", "email", "paid_search"),
  ts = as.POSIXct("2026-01-01", tz = "UTC") + c(0, 1, 2, 3) * 86400,
  conv = c(0, 0, 0, 1)
)
paths <- build_paths(journey, id = "cust", channel = "channel",
                     timestamp = "ts", conversion = "conv")

td <- credit_time_decay(paths, decay = 0.5)
td[, c("channel", "time_to_conversion", "credit")]
```

The touch at the moment of conversion gets the most, the one a day earlier
half as much, and so on backward. Same kernel, opposite arrow.

## The same numbers

Line them up. Normalise the adstock kernel so both sum to 1, and reverse the
credit vector so both run from "closest to the event" outward:

```{r}
forward <- adstock_weights(max_lag = 4, decay = 0.5)
backward <- rev(td$credit)

round(rbind(forward = forward, backward = backward), 6)
all.equal(forward, backward)
```

Identical. Not analogous — identical.

The reason is that both are the geometric kernel $\theta^{d}$ evaluated at a
distance $d$ from an event, normalised over the periods in scope. Adstock
measures $d$ as *time since the spend*, running forward. Time-decay credit
measures $d$ as *time until the conversion*, running backward. Change the sign
of the distance and one becomes the other.

$$
\text{adstock}: \quad w_i \propto \theta^{\,t_i - t_{\text{spend}}}
\qquad\qquad
\text{credit}: \quad w_i \propto \theta^{\,t_{\text{conv}} - t_i}
$$

## Why this is in the API rather than a footnote

Because it means there is one vocabulary rather than two, and practitioners
already have that vocabulary: they say "television works for about three
weeks", not "television has a decay coefficient of 0.79".

```{r}
theta <- decay_from_half_life(3)
theta

# Forward
round(adstock_geometric(c(1000, 0, 0, 0, 0), decay = theta), 2)

# Backward, same argument, same meaning
round(credit_time_decay(paths, decay = theta)$credit, 4)
```

`decay_from_half_life()`, `half_life()` and `effective_window()` serve both
halves of the package because there is only one kernel to describe. A media
planner who has decided television's half-life is three weeks has, without
doing anything else, also specified a defensible time-decay attribution window
— and `effective_window()` will tell them it covers ten periods, which is a far
better basis for a lookback window than reaching for thirty days.

## What the symmetry does not give you

It is a symmetry of *arithmetic*, not of evidence.

Adstock is a claim about a causal mechanism: money spent in week one still
moves sales in week three, and a regression on adstocked spend estimates how
much. It can be wrong, and it is testable — the transform either improves
out-of-sample prediction of the KPI or it does not, which is exactly what
`tune_carryover()` measures.

Time-decay credit is a claim about nothing. It is a convention for dividing an
observed conversion among observed touchpoints. Recency is a reasonable basis
for that convention and there is no experiment that could confirm it, because
the quantity it estimates does not exist independently of the rule. Running the
same kernel backward does not import adstock's evidential status into
attribution.

Keeping both halves in one package makes the shared arithmetic visible, and
that is worth doing. It should not be allowed to blur the difference between a
mechanism and a bookkeeping choice.

## Where the symmetry stops

It holds exactly for the geometric kernel on evenly spaced touches. It stops
being exact in three ordinary situations, all of which are the attribution
side's problem rather than the transform side's:

- **Uneven spacing.** Adstock operates on a regular series by construction.
  Touchpoints arrive whenever they arrive, so `credit_time_decay()` evaluates
  the kernel at the actual elapsed time rather than at an integer lag.
- **Truncation.** Adstock's kernel is normalised over `max_lag` periods; credit
  is normalised over however many touches a journey happens to contain.
- **Non-geometric kernels.** `adstock_weibull()` has no attribution
  counterpart in this package. It could have one — `credit_custom()` takes an
  arbitrary function of rank, journey length and recency — but a delayed-peak
  credit rule would be asserting that touches from a fortnight ago deserve more
  credit than yesterday's, and that is a strange thing to assert without
  evidence.

```{r}
# A Weibull-shaped credit rule, if you wanted one. The peak falls two days
# before the conversion, so the closing touch gets almost nothing.
weibull_credit <- credit_custom(paths, function(rank, n, recency) {
  dweibull(recency + 0.5, shape = 3, scale = 2.5)
})
weibull_credit[, c("channel", "time_to_conversion", "credit")]
```

The hook is there, and it does what a delayed-peak kernel should: the touch two
days out receives the most credit and the one at the moment of conversion
receives 5%. Whether that number means anything is a separate question, and the
honest answer is usually no — a rule asserting that a touch from two days ago
deserves nine times the credit of the click that closed the sale is a claim
about causation, made without evidence, dressed as a convention.
