---
title: "Tuning with tidymodels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tuning with tidymodels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)

has_recipes <- requireNamespace("recipes", quietly = TRUE)
has_tune <- all(vapply(
  c("recipes", "rsample", "parsnip", "tune", "dials", "workflows",
    "yardstick", "glmnet"),
  requireNamespace, logical(1), quietly = TRUE))
knitr::opts_chunk$set(eval = has_recipes)
```

```{r eval = !has_recipes, echo = FALSE, comment = NA}
cat("The recipes package is not installed, so this vignette's code was not run.\n")
```

Carryover decay, saturation shape and a model's own penalty are not
independent. A longer carryover shifts the media series' scale, which changes
which penalty is optimal, which changes which saturation point best fits the
residual curvature. Choosing them one at a time — decay by cross-validation,
saturation by eye, penalty by `glmnet`'s own path — finds a corner of the
parameter space and calls it an optimum.

Tuning them jointly is what the recipes layer is for. It is the entire reason
these transforms are recipe steps rather than functions you call beforehand.

```{r message = FALSE}
library(mediamix)
library(recipes)
data(mm_weekly)

channels <- c("tv", "video", "search", "social", "display")
```

## The problem the step solves

Adstock is a filter with memory. Prepare a recipe on the first two years and
bake it on the third, and the first rows of the third year have no history.

`recipes::step_lag()` answers this by emitting `NA`, silently. Its documented
remedy, `step_naomit()`, changes the row count — and a step that changes the
row count breaks `tune_grid()` outright, because the resampling machinery needs
assessment predictions to line up with assessment rows. That is not an escape
hatch; it is a dead end.

`step_adstock()` stores *filter state* instead of rows:

```{r}
train <- mm_weekly[mm_weekly$date < as.Date("2025-07-01"), ]
test <- mm_weekly[mm_weekly$date >= as.Date("2025-07-01"), ]

rec <- recipe(revenue ~ ., data = train) |>
  step_adstock(all_of(channels), index = "date", by = "geo", decay = 0.7) |>
  prep()

tidy(rec, number = 1)
```

One number per column per geography. Not lag-1 rows of data — one double. The
geometric filter is an infinite impulse response whose entire memory is
$a_{t-1}$, so that single number *is* the history. No training observations are
retained, which sidesteps both the object-size objection and the privacy one.

Baking on contiguous new data warm-starts from it:

```{r}
baked <- bake(rec, new_data = test)
c(rows_in = nrow(test), rows_out = nrow(baked),
  missing = sum(is.na(baked$tv)))
```

Row count preserved, order preserved, no missing values at the boundary. And
the result is exactly what you would have got by filtering the whole series at
once:

```{r}
full <- mm_weekly[order(mm_weekly$geo, mm_weekly$date), ]
full$expected <- ave(full$tv, full$geo,
                     FUN = function(z) adstock_geometric(z, decay = 0.7))

check <- merge(baked[, c("date", "geo", "tv")],
               full[, c("date", "geo", "expected")],
               by = c("date", "geo"))
all.equal(check$tv, check$expected)
```

That equality is the step's whole contract, and it is enforced by the test
suite.

## What happens when the data does not continue

`bake()` compares the start of the new data against what `prep()` saw and acts
accordingly. Four cases, all explicit:

```{r}
# Overlapping the training period: cold start, so re-baking the training data
# reproduces the prepared output rather than double-counting its carryover.
all.equal(bake(rec, new_data = train)$tv, juice(rec)$tv)
```

```{r}
# A gap: cold start, and say so.
gapped <- test[test$date >= as.Date("2025-10-01"), ]
invisible(tryCatch(bake(rec, new_data = gapped),
                   warning = function(w) message(conditionMessage(w))))
```

An unseen group warns the same way. `carry_over = "warm"` turns those warnings
into errors when you want the pipeline to insist on contiguity;
`carry_over = "cold"` disables state entirely and reproduces lag-style
behaviour.

Two things worth noting. The step is strictly causal, unlike
`recipes::step_window()` and `recipes::step_impute_roll()`, which are centred
and fill a series' leading edge using future values — applied to a media filter
that leaks future spend into past adstock and inflates in-sample fit, which is
the exact pathology marketing mix models are accused of. And the step sorts by
`index` internally and restores the original row order before returning, so
nothing downstream ever sees a reordered table.

## Grouping is an argument, not an inherited state

`recipes` has no group awareness — the feature request was closed — and a
`grouped_df` may not survive into `bake()`. So grouping is named explicitly,
following `modeltime`'s precedent:

```{r}
setequal(train$geo, as.character(baked$geo))
tidy(rec, number = 1)$group
```

Each geography carries its own state and its own last-seen index. (`prep()`
converts the character column to a factor along the way, so compare the values
rather than the vectors.)

## Saturation on a comparable scale

Television spend runs to five figures a week here and display to four. An
absolute half-saturation point is therefore a different parameter for every
channel, which makes a single tunable parameter meaningless.

`step_saturation()` expresses it as a *fraction* of each column's own reference
level, learned from the training data:

```{r}
rec2 <- recipe(revenue ~ ., data = train) |>
  step_adstock(all_of(channels), index = "date", by = "geo", decay = 0.7) |>
  step_saturation(all_of(channels), half_max = 0.4, shape = 1.5) |>
  prep()

tidy(rec2, number = 2)
```

`half_max = 0.4` reads the same way for every channel: half the ceiling is
reached at 40% of that channel's peak spend. One parameter now covers all five.

## The tuning parameters

```{r}
carryover_decay()
saturation_shape()
```

```{r}
tunable(rec2$steps[[1]])[, c("name", "source", "component")]
tunable(rec2$steps[[2]])[, c("name", "source", "component")]
```

`carryover_decay()` stops at 0.95 rather than 1 on purpose. A decay of 0.99
implies a half-life of 69 periods, which no ordinary media dataset identifies;
leaving it in the range mostly spends tuning iterations on values that trade
off against the intercept.

## Joint tuning

Now the payoff. One search over carryover, saturation and penalty together,
scored on rolling-origin resampling.

```{r eval = has_tune}
library(rsample)
library(parsnip)
library(workflows)
library(tune)
library(dials)

north <- mm_weekly[mm_weekly$geo == "north", ]
north <- north[order(north$date), ]

folds <- rolling_origin(north, initial = 104, assess = 13, skip = 12,
                        cumulative = TRUE)

tune_rec <- recipe(revenue ~ ., data = north) |>
  # `date` is the filter's index, not a predictor; `geo` is constant within
  # this subset. A recipe does not create dummies by itself, so a leftover
  # character column would reach glmnet's matrix interface and abort.
  update_role(date, new_role = "index") |>
  step_rm(geo) |>
  step_adstock(all_of(channels), index = "date", decay = tune()) |>
  step_saturation(all_of(channels), half_max = tune(), shape = tune()) |>
  step_normalize(all_numeric_predictors())

model <- linear_reg(penalty = tune(), mixture = 0) |>
  set_engine("glmnet")

wf <- workflow() |> add_recipe(tune_rec) |> add_model(model)

params <- extract_parameter_set_dials(wf)
params

set.seed(1)
res <- tune_bayes(wf, resamples = folds, param_info = params,
                  initial = 8, iter = 12,
                  metrics = yardstick::metric_set(yardstick::rmse))

show_best(res, metric = "rmse", n = 5)
```

```{r eval = !has_tune, echo = FALSE, comment = NA}
cat("The full tidymodels stack (rsample, parsnip, workflows, tune, dials,\n",
    "yardstick, glmnet) is not installed here, so the joint-tuning chunk was\n",
    "not run. The recipe it builds is prepped and checked in the package's\n",
    "test suite, so the code above is known to run once those packages are\n",
    "present.\n", sep = "")
```

Four parameters, one search, one resampling scheme, correct train/test
boundaries throughout. Robyn reaches for `nevergrad` — and therefore a Python
runtime — to do the equivalent.

## Leakage, since a reviewer will ask

Under rolling-origin resampling, `prep()` re-runs on each split's analysis set.
The stored state is therefore re-learned from that split's training rows alone
and describes only media that precedes the assessment set. Nothing carries
across splits, and no outcome information crosses any boundary — the state is a
function of past *spend*, which is genuinely known at prediction time.

The same argument covers `step_saturation()`: each column's reference level is
a training-set statistic, re-learned per split, and a test-set spend above it is
not clipped — it simply lands further up the curve.

## If you skip the tuning, skip the step

The recipe steps add machinery: state, a contiguity guard, a boundary contract
to reason about. All of that earns its place because it makes joint tuning
possible. If you are going to fix decay and saturation by hand anyway, use
`media_transform()` on the vectors and keep your pipeline simpler.
