---
title: "Walkthrough: from weekly spend to a budget recommendation"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
```

This article follows one question from raw data to an answer: **if the
marketing budget stays the same, should any of it move between channels?**
It uses the north geography of `mm_weekly`, 156 weeks of five channels,
generated from known parameters so that every step can be checked against the
truth.

The getting-started vignette introduces each function. This article is about
using them together, and about where the answer is solid and where it is not.

```{r}
library(mediamix)
data(mm_weekly)

channels <- c("tv", "video", "search", "social", "display")
truth <- attr(mm_weekly, "truth")
north <- mm_weekly[mm_weekly$geo == "north", ]
north <- north[order(north$date), ]
north$week <- seq_len(nrow(north))
```

## 1. Check the data can answer the question

```{r}
d <- diagnose_media(north, media = channels)
d
d$collinearity[, c("channel", "vif", "most_correlated_with", "correlation")]
```

No channel is collinear enough to make its coefficient meaningless, and every
one varies. Video and display are dark in a third or more of weeks, so they
carry less information than their row count suggests — worth remembering when
their numbers come out.

## 2. Carryover: how well does the data identify it?

Before fitting anything, ask what the data says about each channel's carryover
on its own. The KPI is residualised against the controls first, so the tuning
does not credit trend and seasonality to media:

```{r}
controls <- lm(revenue ~ week + price + seasonality + holiday, data = north)
grid <- seq(0.05, 0.95, by = 0.05)

tuned <- lapply(channels, function(ch) {
  suppressWarnings(tune_carryover(north[[ch]], residuals(controls),
                                  decays = grid))
})
names(tuned) <- channels

rbind(
  truth  = truth$decay[channels],
  best   = vapply(tuned, function(t) t$best$decay, numeric(1)),
  one_se = vapply(tuned, function(t) t$best_1se$decay, numeric(1))
)
```

Television's long carryover is found. For the others, the one-channel-at-a-time
search cannot separate their carryover from zero — `one_se` says the data is
consistent with almost no carryover at all — and `best` lands on values that
are wrong in both directions. That is the expected result of tuning channels
one at a time, as `vignette("carryover")` explains; the remedy is to tune them
jointly inside the model you intend to fit, as `vignette("tidymodels")` shows.

To keep this walkthrough about the *decision* rather than the estimation, the
rest of it uses the generating parameters — the answer a well-specified joint
search is trying to reach. Section 6 comes back to what happens when
television's carryover is uncertain.

## 3. Fit the model

```{r}
spec <- lapply(channels, function(ch) list(
  adstock = list(decay = truth$decay[[ch]]),
  saturation = list(half_max = truth$half_max[[ch]],
                    shape = truth$shape[[ch]])
))
names(spec) <- channels

transform_plan <- function(plan, spec) {
  as.data.frame(lapply(stats::setNames(channels, channels), function(ch) {
    media_transform(plan[[ch]], adstock = spec[[ch]]$adstock,
                    saturation = spec[[ch]]$saturation)
  }))
}

media <- transform_plan(north, spec)
model_data <- cbind(media, north[c("week", "price", "seasonality", "holiday")])
fit <- lm(north$revenue ~ ., data = model_data)
round(rbind(estimate = coef(fit)[channels], truth = truth$beta[channels]))
```

## 4. Where did the revenue come from?

```{r}
contrib <- contributions(media, fit, index = north$date)
avg <- roi(contrib, north[channels])
avg
```

Every channel returns more than it costs on average. On this table alone the
tempting move is to take money from television, the lowest average ROI, and
give it to social, the highest.

## 5. What would the next pound do?

Average ROI describes the budget that was spent. The reallocation question is
about the *margin*, and `marginal_roi()` answers it by re-running the whole
transform on a slightly larger budget:

```{r}
marginal <- do.call(rbind, lapply(channels, function(ch) {
  cbind(channel = ch, marginal_roi(
    north[[ch]], coefficient = coef(fit)[[ch]],
    adstock = spec[[ch]]$adstock, saturation = spec[[ch]]$saturation
  ))
}))
marginal <- merge(marginal[, c("channel", "mroi")],
                  avg[, c("channel", "roi", "spend")], by = "channel")
marginal[order(-marginal$mroi), ]
```

The ranking changes. Social still has the highest marginal return, but
television is no longer at the bottom: its S-shaped response means its next
pound works nearly as hard as its average one. Display is now last, and its
marginal return is below 1 — the last pounds spent on it did not pay for
themselves.

So the candidate move is *display to social*, not *television to social*.

## 6. Test the move before recommending it

A marginal return is a derivative. A real reallocation is a finite change, and
the response curves bend along the way, so simulate the new plan through the
model rather than multiplying by a marginal rate. Everything except the two
channels is held fixed, and the controls do not change, so the difference in
predicted revenue is the difference in media response:

```{r}
media_revenue <- function(plan, spec) {
  m <- transform_plan(plan, spec)
  sum(vapply(channels, function(ch) coef(fit)[[ch]] * sum(m[[ch]]),
             numeric(1)))
}

# Move an amount of budget from one channel to another. Each channel's
# flighting pattern is kept: the change is spread over its weeks in
# proportion to its existing spend.
move <- function(from, to, amount, spec) {
  plan <- north
  plan[[from]] <- plan[[from]] * (1 - amount / sum(plan[[from]]))
  plan[[to]] <- plan[[to]] + amount * plan[[to]] / sum(plan[[to]])
  gain <- media_revenue(plan, spec) - media_revenue(north, spec)
  c(moved = amount, revenue_change = gain, per_pound_moved = gain / amount)
}

tenth <- 0.10 * sum(north$display)
rbind(
  `display -> social` = move("display", "social", tenth, spec),
  `tv -> social`      = move("tv", "social", tenth, spec),
  `display -> social, 2.5x` = move("display", "social", 2.5 * tenth, spec)
)
```

Moving a tenth of display's budget into social raises predicted revenue at no
extra cost. Moving the *same amount* out of television instead — the move the
average-ROI table suggested — gains less, because it gives up television's
still-steep response. Moving two and a half times as much out of display
gains more in total but less per pound, as social climbs its own curve.

Two checks before this goes in a deck. First, is the new plan inside the data?

```{r}
peak <- function(amount) max(north$social + amount * north$social /
                               sum(north$social))
c(observed = max(north$social), smaller_move = peak(tenth),
  larger_move = peak(2.5 * tenth))
```

Both moves push social's busiest week above anything in the data: the smaller
by about 7%, the larger by about 18%. A little beyond the observed range is the
unavoidable price of asking what *more* would do, but the further out a plan
goes, the more of its predicted gain is the fitted curve's shape rather than
evidence. The smaller move stays close enough to the data to be worth
testing; the larger one should wait for that test.

Second, how sensitive is the answer to carryover it could not pin down? In
section 2 the data was consistent with a television decay as low as the
one-SE value. Refit with that and ask again:

```{r}
tv_alt <- tuned$tv$best_1se$decay
spec_alt <- spec
spec_alt$tv$adstock$decay <- tv_alt

media_alt <- transform_plan(north, spec_alt)
fit_alt <- lm(north$revenue ~ .,
              data = cbind(media_alt,
                           north[c("week", "price", "seasonality", "holiday")]))

c(tv_decay = tv_alt,
  tv_mroi_truth = marginal$mroi[marginal$channel == "tv"],
  tv_mroi_alt = marginal_roi(
    north$tv, coefficient = coef(fit_alt)[["tv"]],
    adstock = spec_alt$tv$adstock, saturation = spec_alt$tv$saturation
  )$mroi)
```

Television's marginal return depends heavily on a carryover the data only
loosely identifies: at the shorter decay its next pound no longer pays for
itself. The display-to-social move does not involve television, which is one
more reason to prefer it — it is the recommendation that survives the
uncertainty. Whether television is over- or under-funded is a question this
data cannot settle, and saying so is part of the answer.

## 7. What to say, and what not to

The recommendation this analysis supports is modest and specific: *move
around a tenth of display's budget into social, leave television alone until
its carryover is better measured, and test the move before scaling it.* The last part matters most. Every number
here rests on a model fitted to observational data; a geo experiment — run the
new allocation in one region and compare — turns the model's prediction into
a measurement, and its result can then be used to calibrate the model's
priors or constraints.

What this analysis does not support: a precise revenue forecast for the new
plan, any move that takes a channel far outside the spend it has historically
run at, or a ranking of channels by average ROI.

## Where next

- `vignette("tidymodels")` to tune every channel's carryover and saturation
  jointly rather than one at a time.
- `vignette("carryover")` for why one-at-a-time tuning goes wrong.
- [Methods and references](methodology.html) for the research behind each
  step.
