---
title: "Methods and references"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
options(digits = 4)
library(mediamix)
```

Each function in mediamix implements a published method or a documented
convention. This page says which one, where it comes from, and what choices
the implementation makes, so a result can be reproduced, cited or challenged.

| Function | Method | Source |
|---|---|---|
| `adstock_geometric()` | Geometric (Koyck) carryover | Broadbent (1979); Koyck (1954) |
| `adstock_delayed()` | Delayed-peak geometric carryover | Jin et al. (2017) |
| `adstock_weibull()` | Weibull CDF / PDF carryover | As implemented in Robyn |
| `saturate_hill()` | Hill saturation | Hill (1910); Jin et al. (2017) |
| `saturate_michaelis_menten()`, `saturate_exponential()`, `saturate_power()` | Alternative concave response curves | Standard |
| `tune_carryover()` | Forward-only cross-validation, one-SE rule | Bergmeir & Benítez (2012); Hastie et al. (2009) |
| `tune_carryover_joint()` | Coordinate descent over per-channel carryover in one model | Standard |
| `adstock_steady_state()` | Steady-state filter initialisation | Standard |
| `block_bootstrap()` | Moving block bootstrap | Künsch (1989); Lahiri (2003) |
| `step_adstock()` | Stateful IIR filter across resampling boundaries | This package |
| `contributions()` | Additive decomposition, baseline as residual | Standard |
| `marginal_roi()` | Counterfactual marginal ROI | Jin et al. (2017); Meridian |
| `diagnose_media()` | Variance inflation factors, CV, CPM outliers | Standard |
| `build_paths()` | Journey construction | This package |
| `credit_*()` | Rule-based attribution | Industry conventions |
| `markov_removal()` | First-order Markov removal effects | Anderl et al. (2016) |

## Carryover

**Geometric.** The effect of spend decays by a constant factor
$\theta$ each period, $a_t = x_t + \theta a_{t-1}$ — the Koyck
distributed lag, introduced to advertising as "adstock" by Broadbent (1979).
mediamix defaults to the *normalised* kernel, weights $(1-\theta)\theta^l$
summing to one, as in Jin et al. (2017); Robyn uses the unnormalised recursion.
The two differ only in the scale of the regressor, and so of its coefficient.

**Delayed.** Jin et al. (2017) generalise the geometric kernel to
$w_l = \theta^{(l-\delta)^2}$, whose peak falls $\delta$ periods after the
spend. This is the kernel to use when the response builds before it decays,
and `peak` is read directly in periods:

```{r}
round(rbind(
  geometric = adstock_weights(8, decay = 0.6),
  delayed   = adstock_weights_delayed(8, decay = 0.6, peak = 2)
), 3)
```

**Weibull.** Two parameterisations, both following Robyn's construction: the
`"cdf"` form is a running product of $1 - F(l)$, always decreasing but with
a time-varying rate; the `"pdf"` form uses the density itself and can peak
late. Robyn expresses the scale as a quantile of the window, so its fitted
scale values do not transfer directly.

**Start-up bias.** Every filter here is cold-started at zero unless given
a `state`. For a slow channel that understates the first weeks of the series:
at $\theta = 0.85$ a constant spend reaches 15% of its steady-state adstock
in week one and 90% only after `effective_window(0.85)` weeks. Pass real
pre-period spend through `adstock_state()` when you have it.

## Saturation

The Hill curve $x^s / (x^s + K^s)$ comes from Hill (1910) and is the
saturation curve of Jin et al. (2017) and of most current MMM software.
mediamix parameterises it by `half_max` ($K$, the spend at half the ceiling)
and `shape` ($s$; above 1 gives an S-curve). It is evaluated as
$1 / (1 + (K/x)^s)$, which is algebraically identical but cannot overflow.
Saturation is applied *after* adstock, as in Jin et al.; `media_transform()`
enforces that order and warns if asked to reverse it.

## Choosing carryover by cross-validation

`tune_carryover()` scores each candidate decay by out-of-sample prediction of
the KPI, on forward-only splits — rolling-origin evaluation, which Bergmeir and
Benítez (2012) recommend for time series because it never trains on the
future. Because adstock is causal, the series is transformed once and then
split; the package's tests enforce the causality this relies on.

The score is pooled across splits by default: every out-of-sample forecast is
collected and the metric computed once, so RMSE is a true root mean squared
forecast error.

Carryover is usually weakly identified, so a single minimum overstates what the
data knows. `best_1se` applies the one-standard-error rule (Breiman et al.,
1984; Hastie et al., 2009, §7.10), choosing the shortest carryover the
resampling cannot distinguish from the best. mediamix uses the *paired* form:
candidates are scored on the same splits, so a candidate is compared with the
best by the standard error of their per-split difference, not of either one's
mean. The unpaired form is dominated by how hard each period is to forecast,
which is shared by every candidate, and admits almost the whole grid.

**Several channels.** Tuned one at a time, a channel's carryover absorbs the
other channels' effects and the controls'. `tune_carryover_joint()` puts every
channel and the controls into one model and chooses each channel's
`(max_lag, decay)` by coordinate descent: each channel in turn is set to its
best grid point with the rest held fixed, repeating until a sweep changes
nothing. Every accepted step lowers the cross-validated error, so it
terminates, at a local optimum of the grid; the cost is a few sweeps of the
one-channel grid rather than the full product grid. It reports each channel's
final profile, so a flat one — carryover the data does not identify — is
visible rather than hidden behind a point estimate.

**Start-up bias.** `adstock_steady_state()` returns the filter state that
constant spend at the early-period mean would have left, $m/(1-\theta)$ for
the recursive kernel, so a series does not begin at zero adstock.
`warm_start = TRUE` applies it inside both tuners. It uses media only, never
the KPI, so it cannot leak outcome information into a cross-validation.

## Uncertainty

`block_bootstrap()` implements the moving block bootstrap of Künsch (1989):
resample overlapping blocks of consecutive rows, concatenate them to the
original length, recompute the statistic, and report percentile intervals.
Blocks preserve the autocorrelation that carryover creates, which resampling
single weeks would destroy and so understate uncertainty. The default block
length $\lceil n^{1/3} \rceil$ is the usual rate-optimal order (Lahiri,
2003); for media data a block at least as long as the carryover's effective
window is a sensible floor. With `by`, blocks are drawn within each group.

The transform should be computed once, on the original series, before
bootstrapping: re-adstocking resampled rows would carry spend across the joins
between blocks. The resulting intervals describe sampling uncertainty given
the transform parameters, not the uncertainty in choosing them.

## Joint tuning with tidymodels

`step_adstock()` stores the geometric filter's single-number state at `prep()`
and warm-starts `bake()` from it when the new data continues the training
series. That is what lets carryover, saturation and a model penalty be tuned
jointly by `tune`'s resampling, with no rows lost at the boundary. The
contiguity check classifies new data as contiguous, overlapping, gapped or
unseen, and only warm-starts in the first case.

## Contributions and return on investment

`contributions()` computes $\beta_j z_{jt}$ for each channel and takes the
baseline as the fitted value minus total media, so the decomposition sums
exactly to the model's prediction. For an additive model this is also each
channel's removal effect — what the model predicts would be lost with that
channel at zero.

`marginal_roi()` follows the counterfactual definition used by Jin et al.
(2017) and by Google's Meridian: increase spend by a small proportion,
re-run carryover and saturation, and divide the extra response by the extra
spend. Differentiating the saturation curve at the mean adstocked level instead
is a common shortcut and has two problems. Multiplying by the kernel's first
weight to get back to spend counts only the period of spend and drops every
later period's carryover, and the slope at the mean is not the mean of the
slope. The simulation has neither problem.

```{r}
set.seed(3)
x <- pmax(0, rnorm(200, 1000, 300))
z <- adstock_geometric(x, decay = 0.85)
c(shortcut  = mroi(mean(z), 1000, half_max = 1000) * (1 - 0.85),
  simulated = marginal_roi(x, 1000, adstock = list(decay = 0.85),
                           saturation = list(half_max = 1000))$mroi)
```

## Diagnostics

`diagnose_media()` reports variance inflation factors (regressing each channel
on the others), the coefficient of variation, the share of dark periods and
implied cost-per-mille outliers (a median-absolute-deviation rule). With
`decay`, collinearity is measured on adstocked media, which is what the model
actually sees. With `by`,
each is computed within series and the worst case reported, since pooling
geographies that differ in scale manufactures correlation.

## Journeys and attribution

`build_paths()` has no single literature source; it encodes the construction
choices — splitting on conversion and inactivity, lookback, null paths,
direct traffic, repeats, tie-breaking — that the attribution literature
assumes have already been made. The rule-based credits (`credit_first()`,
`credit_last()`, `credit_linear()`, `credit_position()`,
`credit_time_decay()`) are industry conventions and make no causal claim.

`markov_removal()` implements the first-order Markov model of Anderl et al.
(2016). Journeys are walks from a start state through channels to one of two
absorbing states. The conversion probability comes from the fundamental matrix,
$(I - Q)^{-1} R$, so it is exact for the fitted chain rather than simulated. A
channel's removal effect is the proportional drop in that probability when
every transition into the channel is redirected to the null state, and
conversions are divided in proportion to removal effects.

```{r}
ev <- data.frame(id = c("j1", "j1", "j2", "j3"), ch = c("a", "b", "a", "b"),
                 ts = c(1, 2, 1, 1), conv = c(0, 1, 0, 1))
markov_removal(build_paths(ev, id = "id", channel = "ch", timestamp = "ts",
                           conversion = "conv"))
```

Here the conversion probability is 2/3. Removing `a` halves it and removing `b`
makes conversion impossible, so `b` gets twice `a`'s credit.

## What none of this is

None of these methods estimates incrementality from observational data alone.
Regression-based MMM is identified only as well as its specification and the
variation in the data allow, and attribution rules — including Markov removal
effects — divide observed conversions by a convention. Randomised experiments
(geo experiments for media, holdouts for digital channels) measure what these
methods estimate, and are the standard way to calibrate them.

## References

Anderl, E., Becker, I., von Wangenheim, F. and Schumann, J. H. (2016). Mapping
the customer journey: Lessons learned from graph-based online attribution
modeling. *International Journal of Research in Marketing*, 33(3), 457–474.
<https://doi.org/10.1016/j.ijresmar.2016.03.001>

Bergmeir, C. and Benítez, J. M. (2012). On the use of cross-validation for
time series predictor evaluation. *Information Sciences*, 191, 192–213.

Breiman, L., Friedman, J., Olshen, R. and Stone, C. (1984). *Classification and
Regression Trees*. Wadsworth.

Broadbent, S. (1979). One way TV advertisements work. *Journal of the Market
Research Society*, 21(3), 139–166.

Hastie, T., Tibshirani, R. and Friedman, J. (2009). *The Elements of
Statistical Learning* (2nd ed.). Springer.

Hill, A. V. (1910). The possible effects of the aggregation of the molecules of
haemoglobin on its dissociation curves. *The Journal of Physiology*, 40,
iv–vii.

Jin, Y., Wang, Y., Sun, Y., Chan, D. and Koehler, J. (2017). *Bayesian methods
for media mix modeling with carryover and shape effects.* Google Inc.
<https://research.google/pubs/pub46001/>

Koyck, L. M. (1954). *Distributed Lags and Investment Analysis.*
North-Holland.

Künsch, H. R. (1989). The jackknife and the bootstrap for general stationary
observations. *The Annals of Statistics*, 17(3), 1217–1241.

Lahiri, S. N. (2003). *Resampling Methods for Dependent Data.* Springer.
